An Examination of Sport Climbing Competition Format and Scoring System

### 2021-07-30 manuscript/Rmd/draft.Rmd
More edits.

These three different climbing disciplines demand different sets of skills and, often, athletes specialize in a single event. However, since only one set of Olympic medals is awarded to sport climbing, rather than choosing only one of these disciplines to include in the Olympics, all three events were chosen to be included as a sort of climbing triathlon.

The decision to combine the three climbing events and only award one set of medals for both men's and women's events in the Olympics has received a large amount of criticism from climbing athletes all over the world (CITATION NEEDED). In a series of interviews conducted by Climbing Magazine in 2016, a number of climbers shared their thoughts and concerns about the new Olympics climbing format. Climber Lynn Hill compared the idea of combining speed climbing, bouldering, and lead climbing to "asking a middle distance runner to compete in the sprint". She then added "Speed climbing is a sport within our sport". Other climbers also hold the same opinion as Hill regarding speed climbing, using words and phrases like "bogus", "a bummer", "less than ideal", "not in support", and "cheesy and unfair" to describe the new combined competition format. Courtney Woods stated "Speed climbers will have the biggest disadvantage because their realm isn’t based on difficult movements". Mike Doyle believed "Honestly, the people that will suffer the most are the ones that focus only on speed climbing. Those skills/abilities don’t transfer as well to the other disciplines". Most climbers also expressed their hope for a change in the format in future tournaments, with some calling for each discipline to have its own set of medals. See @blanchard2016 for more information on these interviews.

In terms of participation and scoring, both sport climbing competitions for male and female at the 2020 Summer Olympics begin with 20 climbers who has previously qualified for the Olympics from qualifying events held in 2019 and 2020. All 20 athletes compete in each of the three disciplines in the qualification round, and their performances in each concentration are ranked from 1 to 20. A competitor's combined score is computed as the product of their ranks in the three events; specifically, 
\begin{equation}
Score_i = R^S_i\times R^B_i\times R^L_i,
\end{equation} 
where $R^S_i$, $R^B_i$, and $R^L_i$ are the ranks of the $i$-th competitor in speed climbing, bouldering, and lead climbing, respectively.

The top 8 lowest-scoring climbers in the qualification round advance to the finals where they once again compete in all three of speed climbing, bouldering, and lead climbing. The total score for each person in the final stage is determined by multiplying their ranks in each discipline, similar to the qualification round. This implies that the climber with the lowest product of ranks in the final wins the gold medal. This type of scoring system heavily rewards high finishes and relatively ignores poor finishes.  For instance, if climber A finished 1st, 20th, and 20th and climber B finished 10th, 10th, and 10th, climber B would have a score of 1000 whereas climber A would have a much better score of 400, despite finishing last in 2 out of 3 of the events.

>>>>

[//]: (I think this paragraph and the next should get moved somewhere else (or removed)  Maybe in the conclusions?  Definitely for the blog post include this, but I'm not sure how much of this needs to be in the actual article)
These three different climbing disciplines demand different sets of skills and, often, athletes specialize in a single event. However, since only one set of Olympic medals is awarded to sport climbing, rather than choosing only one of these disciplines to include in the Olympics, all three events were chosen to be included as a sort of climbing triathlon.

The decision to combine the three climbing events and only award one set of medals for both men's and women's events in the Olympics has received a large amount of criticism from climbing athletes all over the world (CITATION NEEDED). In a series of interviews conducted by Climbing Magazine in 2016, a number of climbers shared their thoughts and concerns about the new Olympics climbing format. Climber Lynn Hill compared the idea of combining speed climbing, bouldering, and lead climbing to "asking a middle distance runner to compete in the sprint" (CITATION NEEDED). She then added "Speed climbing is a sport within our sport". Other climbers also hold the same opinion as Hill regarding speed climbing, using words and phrases like "bogus", "a bummer", "less than ideal", "not in support", and "cheesy and unfair" to describe the new combined competition format. Courtney Woods stated "Speed climbers will have the biggest disadvantage because their realm isn’t based on difficult movements". Mike Doyle believed "Honestly, the people that will suffer the most are the ones that focus only on speed climbing. Those skills/abilities don’t transfer as well to the other disciplines". Most climbers also expressed their hope for a change in the format in future tournaments, with some calling for each discipline to have its own set of medals. See @blanchard2016 for more information on these interviews.

At the Tokyo Olympics, both sport climbing competitions for male and female at the 2020 Summer Olympics begin with 20 climbers who has previously qualified for the Olympics from qualifying events held in 2019 and 2020. All 20 athletes compete in each of the three disciplines in the qualification round, and their performances in each concentration are ranked from 1 to 20. A competitor's combined score is computed as the product of their ranks in the three events; specifically, 
\begin{equation}
Score_i = R^S_i\times R^B_i\times R^L_i,
\end{equation} 
where $R^S_i$, $R^B_i$, and $R^L_i$ are the ranks of the $i$-th competitor in speed climbing, bouldering, and lead climbing, respectively.

The 8 climbers with the lowest score (i.e. lowest product of ranks across the three discplines) in the qualification round advance to the finals where they once again compete in all three disciplines. The final score for each person in the final stage is determined by multiplying their ranks in each discipline, similar to the qualification round, with the main difference being that there are only 8 competitors in the final as opposed to 20 in the qualification round.  The climbers with the lowest, second lowest, and third lowest product of ranks in the final wins the gold, silver, and bronze medal, respectively.

This type of scoring system heavily rewards high finishes and relatively ignores poor finishes.  For instance, if climber A finished 1st, 20th, and 20th and climber B finished 10th, 10th, and 10th, climber B would have a score of 1000 whereas climber A would have a much better score of 400, despite finishing last in 2 out of 3 of the events.
